---
title: "TidyTuesday — Week 30"
subtitle: "2026-07-26"
author:
  - name: Thomas Reinke
    affiliation:
      - name: Baylor University
        department: Statistics Department
        city: Waco
        state: TX
        country: US
        url: https://www.baylor.edu
    email: thomas_reinke1@baylor.edu
date-format: "MMMM d, yyyy"
format:
  html:
    code-tools:
      source: false
    page-layout: full
    toc-expand: 2
    anchor-sections: false
    citations-hover: true
    footnotes-hover: true
    crossrefs-hover: true
    toc-location: right
    other-links:
      - text: GitHub
        href: https://github.com/thomasqmd
        icon: github
      - text: Website
        href: https://thomasqmd-website.share.connect.posit.cloud/
        icon: globe
      - text: LinkedIn
        href: https://www.linkedin.com/in/thomas-reinke1
        icon: linkedin
theme: [materia, ../../quarto-assets/baylor-html-theme.scss]
code-fold: true
code-overflow: wrap
highlight-style: a11y
toc: true
toc-depth: 5
number-sections: false
number-depth: 1
callout-appearance: minimal
lightbox:
  loop: true
embed-resources: true
editor:
  markdown:
    wrap: 72
---

```{r, setup}
#| include: false
#| message: false
#| eval: true
library(knitr)
library(tidyverse)
library(conflicted)
library(janitor)
library(kableExtra)
library(plotly)
library(patchwork)
library(khroma)
library(sf)
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflicts_prefer(plotly::layout)
knitr::opts_chunk$set(
    comment = "#>",
    message = FALSE,
    warning = FALSE,
    cache = FALSE,
    echo = TRUE,
    eval = TRUE,
    tidy.opts = list(width.cutoff = 100),
    tidy = FALSE,
    fig.align = "center"
)
ggplot2::theme_set(ggplot2::theme_minimal())
ggplot2::theme_update(panel.grid.minor = ggplot2::element_blank())
```

# Near-Death Experience Classifications

## Data

Load the week's data from the `data/` folder:

```{r load-data}
data <- read_csv("data/nde_experiences.csv")
glimpse(data)
```

## Analysis

This week's data contains 590 near-death experience (NDE) narratives collected from online databases, along with AI-derived classification flags and Greyson NDE Scale scores.

### Classification and Thematic Elements by Gender

The near-death experience dataset includes 589 narratives (339 female, 250 male). Beyond the primary classification of near-death experiences (NDE, Probable NDE, Possible NDE, NDE-Like, FDE, SDE, SMR), the collection includes AI-extracted flags indicating specific experiential themes.

The visualizations below examine gender distributions across both narrative classifications and thematic elements:

```{r vis-gender-breakdown}
#| fig.height: 8
#| fig.width: 10

data_gender <- data |>
    filter(!is.na(gender)) |>
    mutate(
        gender = factor(gender, levels = c("M", "F"), labels = c("Male", "Female")),
        classification = factor(classification, levels = c(
            "NDE", "Probable NDE", "Possible NDE", "NDE-Like", "NDE-Like;NDE", "FDE", "SDE", "SMR"
        ))
    )

# Panel 1: Classification Counts and Category Shares
class_df <- data_gender |>
    count(classification, gender) |>
    group_by(classification) |>
    mutate(total = sum(n), pct = n / total * 100) |>
    ungroup()

p_class <- class_df |>
    ggplot(aes(
        x = classification, y = n, fill = gender,
        text = paste0(
            "<b>Classification:</b> ", classification,
            "<br><b>Gender:</b> ", gender,
            "<br><b>Narrative Count:</b> ", n,
            "<br><b>Category Share:</b> ", round(pct, 1), "%"
        )
    )) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
    khroma::scale_fill_vibrant() +
    labs(
        title = "NDE Classification Counts by Gender",
        subtitle = "Breakdown of reported experience types across Male (N=250) and Female (N=339)",
        x = "Classification Category", y = "Number of Narratives", fill = "Gender"
    ) +
    theme_minimal(base_size = 12) +
    theme(
        plot.title = element_text(face = "bold", size = 13),
        axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "top"
    )

# Panel 2: Core NDE Thematic Element Counts
ai_df <- data_gender |>
    pivot_longer(cols = starts_with("ai_"), names_to = "feature", values_to = "present") |>
    group_by(gender, feature) |>
    summarise(
        count = sum(present, na.rm = TRUE),
        total_gender = n(),
        pct = count / total_gender * 100,
        .groups = "drop"
    ) |>
    mutate(
        feature_clean = case_when(
            feature == "ai_clinical" ~ "Clinical Death",
            feature == "ai_obe" ~ "Out-of-Body (OBE)",
            feature == "ai_esp" ~ "ESP / Telepathy",
            feature == "ai_unity" ~ "Universal Unity",
            feature == "ai_hellish" ~ "Hellish Experience",
            feature == "ai_world_future" ~ "Future Vision",
            feature == "ai_past_lives" ~ "Past Lives",
            feature == "ai_aliens" ~ "Alien Entities",
            TRUE ~ feature
        )
    )

feat_order <- ai_df |>
    group_by(feature_clean) |>
    summarise(total_cases = sum(count)) |>
    arrange(total_cases) |>
    pull(feature_clean)

ai_df <- ai_df |> mutate(feature_clean = factor(feature_clean, levels = feat_order))

p_feat <- ai_df |>
    ggplot(aes(
        x = feature_clean, y = count, fill = gender,
        text = paste0(
            "<b>Thematic Element:</b> ", feature_clean,
            "<br><b>Gender:</b> ", gender,
            "<br><b>Narrative Count:</b> ", count,
            "<br><b>Prevalence within Gender:</b> ", round(pct, 1), "%"
        )
    )) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
    coord_flip() +
    khroma::scale_fill_vibrant() +
    labs(
        title = "Counts of Core NDE Elements by Gender",
        subtitle = "Number of narratives containing specific experiential features",
        x = NULL, y = "Number of Narratives", fill = "Gender"
    ) +
    theme_minimal(base_size = 12) +
    theme(
        plot.title = element_text(face = "bold", size = 13),
        legend.position = "top"
    )

p_class / p_feat
```


### Geographic Distribution Over Time

The collection contains experiences reported from around the world across several decades (1936–2024). To analyze how reporting expanded geographically, all raw location strings were standardized into ISO3 country codes and tracked by experience year (`exp_date`).

The interactive map below incorporates an **animated timeline** (1950–2024). Press **Play** or drag the slider to observe how reported experiences accumulated across different countries over time.

```{r vis-geographic-animated}
# Standardize location entries
country_clean_map <- c(
    "American Samoa" = "USA", "Argentina" = "ARG", "Australia" = "AUS", "Austria" = "AUT",
    "Belgium" = "BEL", "Brazil" = "BRA", "Bridgeport" = "USA", "Britain" = "GBR",
    "Bulgaria" = "BGR", "Canada" = "CAN", "Costa Rica" = "CRI", "Cuba" = "CUB",
    "Cyprus" = "CYP", "Denmark" = "DNK", "Dominican Republic" = "DOM", "Egypt" = "EGY",
    "England" = "GBR", "France" = "FRA", "Geneva" = "CHE", "Germany" = "DEU",
    "Ghana" = "GHA", "India" = "IND", "Iowa" = "USA", "Iran" = "IRN", "Ireland" = "IRL",
    "Israel" = "ISR", "Italy" = "ITA", "Jerusalem" = "ISR", "Kentucky" = "USA",
    "Macau" = "MAC", "Malta" = "MLT", "Mexico" = "MEX", "Nederland" = "NLD",
    "Netherlands" = "NLD", "New Zealand" = "NZL", "Norway" = "NOR", "Nova Scotia" = "CAN",
    "Ohio" = "USA", "Oregon" = "USA", "Philippines" = "PHL", "Poland" = "POL",
    "polska" = "POL", "Russia" = "RUS", "Saudi Arabia" = "SAU", "South Africa" = "ZAF",
    "South Korea" = "KOR", "Soviet Union" = "RUS", "Spain" = "ESP", "Sweden" = "SWE",
    "Switzerland" = "CHE", "Taiwan" = "TWN", "Tennessee" = "USA", "Trinidad" = "TTO",
    "U" = "USA", "Ukraine" = "UKR", "United Kingdom" = "GBR", "United States" = "USA",
    "Venezuela" = "VEN", "Vietnam" = "VNM", "Wales" = "GBR"
)

iso_to_name <- c(
    "USA" = "United States", "ARG" = "Argentina", "AUS" = "Australia", "AUT" = "Austria",
    "BEL" = "Belgium", "BRA" = "Brazil", "GBR" = "United Kingdom", "BGR" = "Bulgaria",
    "CAN" = "Canada", "CRI" = "Costa Rica", "CUB" = "Cuba", "CYP" = "Cyprus",
    "DNK" = "Denmark", "DOM" = "Dominican Republic", "EGY" = "Egypt", "FRA" = "France",
    "CHE" = "Switzerland", "DEU" = "Germany", "GHA" = "Ghana", "IND" = "India",
    "IRN" = "Iran", "IRL" = "Ireland", "ISR" = "Israel", "ITA" = "Italy",
    "MAC" = "Macau", "MLT" = "Malta", "MEX" = "Mexico", "NLD" = "Netherlands",
    "NZL" = "New Zealand", "NOR" = "Norway", "PHL" = "Philippines", "POL" = "Poland",
    "RUS" = "Russia", "SAU" = "Saudi Arabia", "ZAF" = "South Africa", "KOR" = "South Korea",
    "ESP" = "Spain", "SWE" = "Sweden", "TWN" = "Taiwan", "TTO" = "Trinidad & Tobago",
    "UKR" = "Ukraine", "VEN" = "Venezuela", "VNM" = "Vietnam"
)

df_clean <- data |>
    mutate(
        iso3 = country_clean_map[country],
        exp_year = year(exp_date)
    ) |>
    filter(!is.na(iso3), !is.na(exp_year))

# Build cumulative yearly grid for animation (1950 to 2024)
years_range <- 1950:2024
all_iso3 <- unique(df_clean$iso3)

yearly_counts <- df_clean |>
    filter(exp_year >= 1950) |>
    count(exp_year, iso3, name = "year_n") |>
    complete(exp_year = years_range, iso3 = all_iso3, fill = list(year_n = 0)) |>
    arrange(iso3, exp_year) |>
    group_by(iso3) |>
    mutate(
        cum_n = cumsum(year_n),
        cum_n_log = log10(cum_n + 1),
        country_name = iso_to_name[iso3]
    ) |>
    ungroup() |>
    mutate(
        hover_text = paste0(
            "<b>", country_name, " (", iso3, ")</b>",
            "<br><b>Year:</b> ", exp_year,
            "<br><b>New Reports in Year:</b> ", year_n,
            "<br><b>Cumulative Reports (1950–", exp_year, "):</b> ", cum_n
        )
    )

max_cum_log <- max(yearly_counts$cum_n_log)
tick_vals <- c(0, 1, 5, 20, 50, 150, 350)

# Convert Paul Tol sequential palette from khroma to Plotly colorscale format
pt_colorscale <- function(palette_name = "YlOrBr", n = 9) {
    cols <- as.character(khroma::color(palette_name)(n))
    seqs <- seq(0, 1, length.out = n)
    lapply(seq_along(seqs), function(i) list(seqs[i], cols[i]))
}

# Animated choropleth map with log-scaled Paul Tol color scale
plot_ly(
    data = yearly_counts,
    type = "choropleth",
    locations = ~iso3,
    z = ~cum_n_log,
    zmin = 0,
    zmax = max_cum_log,
    text = ~hover_text,
    frame = ~exp_year,
    colorscale = pt_colorscale("YlOrBr"),
    colorbar = list(
        title = "Cumulative<br>Reports",
        tickvals = log10(tick_vals + 1),
        ticktext = as.character(tick_vals)
    ),
    hoverinfo = "text"
) |>
    layout(
        title = list(
            text = "<b>Global Spatio-Temporal Distribution of NDE Reports (1950–2024)</b>",
            font = list(size = 15, family = "sans-serif")
        ),
        geo = list(
            showframe = FALSE,
            showcoastlines = TRUE,
            coastlinecolor = toRGB("grey70"),
            projection = list(type = "natural earth"),
            bgcolor = "white",
            landcolor = toRGB("grey95")
        ),
        margin = list(t = 60, b = 40, l = 20, r = 20)
    ) |>
    animation_opts(
        frame = 250,
        transition = 100,
        redraw = TRUE
    ) |>
    animation_slider(
        currentvalue = list(prefix = "Year of Experience: ", font = list(color = "#2b5c8f", size = 14))
    )
```


