---
title: "TidyTuesday — Week 30"
subtitle: "2026-07-26"
author:
  - name: Thomas Reinke
    affiliation:
      - name: Baylor University
        department: Statistics Department
        city: Waco
        state: TX
        country: US
        url: https://www.baylor.edu
    email: thomas_reinke1@baylor.edu
date-format: "MMMM d, yyyy"
format:
  html:
    code-tools:
      source: false
    page-layout: full
    toc-expand: 2
    anchor-sections: false
    citations-hover: true
    footnotes-hover: true
    crossrefs-hover: true
    toc-location: right
    other-links:
      - text: GitHub
        href: https://github.com/thomasqmd
        icon: github
      - text: Website
        href: https://thomasqmd-website.share.connect.posit.cloud/
        icon: globe
      - text: LinkedIn
        href: https://www.linkedin.com/in/thomas-reinke1
        icon: linkedin
theme: [materia, ../../quarto-assets/baylor-html-theme.scss]
code-fold: true
code-overflow: wrap
highlight-style: a11y
toc: true
toc-depth: 5
number-sections: false
number-depth: 1
callout-appearance: minimal
lightbox:
  loop: true
embed-resources: true
editor:
  markdown:
    wrap: 72
---

```{r, setup}
#| include: false
#| message: false
#| eval: true
library(knitr)
library(tidyverse)
library(conflicted)
library(janitor)
library(kableExtra)
library(plotly)
library(patchwork)
library(khroma)
library(sf)
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflicts_prefer(plotly::layout)
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo = TRUE,
  eval = TRUE,
  tidy.opts = list(width.cutoff = 100),
  tidy = FALSE,
  fig.align = "center"
)
ggplot2::theme_set(ggplot2::theme_minimal())
ggplot2::theme_update(panel.grid.minor = ggplot2::element_blank())
```

# Near-Death Experience Classifications

## Data

Load the week's data from the `data/` folder:

```{r load-data}
data <- read_csv("data/nde_experiences.csv")
glimpse(data)
```

## Analysis

This week's data contains 590 near-death experience (NDE) narratives collected from online databases, along with AI-derived classification flags and Greyson NDE Scale scores.

### Classification by Gender

The `classification` column reveals that not all narratives in this collection are actually near-death experiences. Some are classified as FDE (far-death experience), SDE (near-death experience without clinical death), or SMR (survivor's guilt memory). The strip plot below shows how these classifications distribute across genders.

```{r vis-classification-strip}
strip <- data |>
  mutate(
    gender = factor(gender, levels = c("M", "F"), labels = c("Male", "Female")),
    classification = factor(classification, levels = c(
      "NDE", "Probable NDE", "Possible NDE", "NDE-Like", "NDE-Like;NDE",
      "FDE", "SDE", "SMR"
    ))
  ) |>
  ggplot(aes(x = classification, y = gender, color = classification,
             text = paste0("Classification: ", classification, "<br>Gender: ", gender))) +
  geom_jitter(height = 0.15, width = 0.1, alpha = 0.7, size = 2) +
  scale_color_muted() +
  labs(
    x = "Classification",
    y = "Gender",
    title = "NDE Classifications by Gender",
    subtitle = "Each point represents one narrative (N = 589)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    legend.position = "none",
    plot.margin = margin(20, 20, 30, 20)
  )

ggplotly(strip, tooltip = "text")
```

### Geographic Distribution

The vast majority of narratives come from the United States, but experiences are reported from over 20 countries. This map shows where these experiences were reported.

```{r vis-geographic}
# Prepare country mapping
country_data <- data |>
  count(country, sort = TRUE) |>
  rename(region = country) |>
  filter(!is.na(region), region != "unknown") |>
  mutate(n = as.numeric(n))

# Map country names to ISO3 codes for plot_ly
iso_map <- tibble(
  country = c("United States", "Canada", "Australia", "France", "England",
              "Belgium", "New Zealand", "United Kingdom", "Mexico", "Argentina",
              "Germany", "India", "Sweden", "Switzerland", "Poland",
              "Britain", "Ireland", "Norway", "Netherlands", "Italy",
              "Spain", "Portugal", "Japan", "South Africa", "Brazil",
              "Russia", "China", "Egypt", "Turkey", "Kenya", "Colombia",
              "Chile", "Venezuela", "Peru", "Greece", "Czech Republic",
              "Denmark", "Finland", "Israel", "Lebanon", "Malaysia",
              "Philippines", "Singapore", "Taiwan", "Thailand", "UAE",
              "Uganda", "Zimbabwe"),
  iso3 = c("USA", "CAN", "AUS", "FRA", NA,
           "BEL", "NZL", "GBR", "MEX", "ARG",
           "DEU", "IND", "SWE", "CHE", "POL",
           NA, "IRL", "NOR", "NLD", "ITA",
           "ESP", "PRT", "JPN", "ZAF", "BRA",
           "RUS", "CHN", "EGY", "TUR", "KEN", "COL",
           "CHL", "VEN", "PER", "GRC", "CZE",
           "DNK", "FIN", "ISR", "LBN", "MYS",
           "PHL", "SGP", "TWN", "THA", "ARE",
           "UGA", "ZWE")
)

# Merge with country counts
country_counts <- iso_map |>
  left_join(country_data, by = c("country" = "region")) |>
  filter(!is.na(iso3), !is.na(n))

# Create choropleth
plot_ly(
  type = "choropleth",
  locations = country_counts$iso3,
  z = country_counts$n,
  text = paste0("<b>", country_counts$country, "</b><br>Count: ", country_counts$n),
  colorscale = list(
    c(0, "#FFFFCC"),
    c(0.2, "#CCEBC5"),
    c(0.4, "#78C679"),
    c(0.6, "#41AB5D"),
    c(0.8, "#238443"),
    c(1.0, "#00441B")
  ),
  colorbar = list(title = "Count"),
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = "Geographic Distribution of NDE Narratives",
      font = list(size = 15, family = "sans-serif")
    ),
    subtitle = list(
      text = "Experiences reported by country (N = 589)",
      font = list(size = 11, color = rgb(0.25, 0.25, 0.25))
    ),
    geo = list(
      showframe = FALSE,
      showcoastlines = TRUE,
      coastlinecolor = "grey80",
      projection = list(type = "natural earth"),
      bgcolor = "white",
      landcolor = "grey92"
    ),
    margin = list(t = 80, b = 40, l = 40, r = 40)
  )
```

